Unicode Implementation Issues \thanks { This is a working document for the Rexx Language Association (RexxLA). There will be one or more separate requirements documents. It is written in \LaTeX\ \fmtversion\ and was rendered \today\ using \enginedetails. }

### 2023-10-04 ARB/standards/work-in-progress/unicode/Unicode.Implimentation.tex
Add files via upload
Externalize bibliography for Unicode.tex

%Compile with latexmk -lualatex
%
% Changes
% Table of contents
% \underline{default}
% Split nomenclature section
% Link to 3.9 Unicode Encoding Forms
% Codepoints/EGCs option in support of charin()
% Correct and expand constants example
% Index
% stream - seek and update
% TUTOR
% Reference ARB GitHub issues
%
% \documentclass[draft]{article}
\documentclass[b4paper]{article}
\usepackage{enumitem}
\usepackage{eqparbox}
\usepackage{expl3}
\usepackage{fontspec}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{listings}         %Requires luatex engine to handle UTF-8
                              %Consider using minted, but ...
\usepackage{makeidx}
\usepackage{upquote}
\usepackage{cleveref}

\makeindex
\setmainfont{CMU Serif}
% texlive-cm-unicode or retrieve cm-unicode with MikTex Console

@@

@@

\section{Background}
The Architecture Review board is investigating areas in which Rexx may be
in need of enhancements, some of which are listed in \cite{Req}.
Many modern languages support Unicode encoded as
\href{https://datatracker.ietf.org/doc/rfc3629/}{UTF-8} (\cite{RFC 3629}),
and the \href{https://www.ietf.org/}{IETF}, in \cite{RFC 5198},
has mandated the use of UTF-8 with NFC in new protocols.
Rexx will need to support UTF-8 if it is not to become a backwater,
and that support should not break existing code.
Elaboration of this issue may be added to \cite{Req} \cref{issue:MB,issue:UTF-8}.

\section{Scope}
This document only describes issues related to \cref{issue:EA,issue:MB}, expanded alphabet and
multi-octet representations of characters in Unicode \cref{issue:MB,issue:UTF-8}; it does not
address such issues as bidirectional text, nor does it address legacy
DBCS support. It presents suggested solutions for requirements that
are presented in other documents.  Also, most of the details are under
active discussion, so everything here is provisional.

@@

The definitions given in \cite{Glossary} and \cite{Unicode} take
precedence over those given here. Quoted text is taken from those sources.
Except for definitions taken from official IETF and Unicode documents,
the nomenclature used here is illustrative rather than normative;
language design teams will formally define, e.g., method names, encoding of parameters.
Most of the sections assume that there will be destinct string types
for extended grapheme clusters (\cref{issue:NT,issue:c2u}), Unicode code points (\cref{issue:NT,issue:c2u}),
legacy code pages (\cref{issue:Leg,issue:mLS,issue:iconv})
and raw octets (\cref{issue:ROS,issue:NT,issue:mO}). However, there has been discussion of including
metadata, in which case come of the classes might be merged.

@@

\begin{lstlisting}
    foo = 'Ren!\'{e}!'
    parse upper var foo bar
\end{lstlisting}

@@

\subsection{Identifiers, labels and procedure names}
The current definition restricts symbols to ASCII characters, which
excludes letters used in many languages. If the rules are extended, in
accordance with \cite{Annex31}.  to include Unicode letters and digits
beyond ASCII, the rules for equality of symbols must be addressed.
Are composed and decomposed strings identical? Are base characters
identical to their compatibility alterrnates? Are subscript and
superscript digits distinct from their ASCII counterparts?

When a numeric value is used as a label or procedure name, a precise
definition of the semantics is needed.  While the definition
"Translate Binary to a coded string." of Config\_B2C in ANSI X3J18-199X
is rather vague, it is adequate when the character set is restricted
to an 8-bit character set.  The definition must be replaced by one
that is unamgiguous when applied to Unicode.  (\ref{issue:IAF})
There are several
obvious options.
\begin{enumerate}
\item Treat every 32 bits as a Unicode scalar
\item Treat every 8 bits as part of a UTF-8 sequence
\item Treat every 8 bits as a character in a legacy code page.
\end{enumerate}
Each of these options has disadvantages.
The same issue exists for Config\_C2B.

@@

@@

\subsection{Unassigned and invalid code points}
There has been some discussion of the need to detect unassigned code points. (\cref{issue:UICP,issue:ED})

@@

\section{ARB Issues}
\label{ARBiss}

This section lists all issues in \cite{Req}, whether or not they relate to Unicode.

\begin{issue}[label=I\arabic*,ref=\fbox{I\arabic*}]
\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730857}%
{Expanded Alphabets}}
\label{issue:EA}

@@

@@

@@

Assuming that the width is in in terms of octets
\begin{lstlisting}[linewidth=35em]
    foo = 'Ren!\'{e}!'
    bar = foo~UTF-8
    /* assumes that width parameter is in octets */
    say foo~c2x(2) /* Displays 00520065006E00E9  */
    say bar~c2x    /* Displays 52656EC3A9        */
\end{lstlisting}

There is an ongoing discussion as to whether an when to allow implicit coercions of type in, e.g.,
\begin{lstlisting}[linewidth=35em]
    foo = .BYTES~new
    foo = 'Ren!\'{e}!'
    !\textellipsis!
    parse var foo ASCII !{\textquotesingle}\'{e}{\textquotesingle}! .
    say ASCII                     /* Displays Ren */
\end{lstlisting}

@@

\subsection{Constants}
Other languages allow specifying Unicode characters using either
the hexadecimal value of the code point or the assigned name of the code point,
(\cref{issue:CUC})
e.g., U+E9
(\cref{issue:NN})
might be coded as {\textbackslash}u\{E9\} or {\textbackslash}u[LATIN SMALL LETTER E WITH ACUTE].
The syntax used for such constants should be consisten across Rexx variants,
including rules for optional spaces between (hex) digits,
and should take into account the recommendations in \cite{RFC 5137}.
Implementations should use the machine readable data bases published by
the Unicode Consortium in order to ease migration to new versions of Unicode.

@@

\begin{tabular}{| l | l | l |}
\hline
Example & Type & Semantics \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}               & CODEPOINTS & Unicode text by code point             \\
                                                           & TEXT       & Unicode text by grapheme cluster (TBD) \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}R                & CODEPOINTS & Unicode text by code point             \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}CODEPOINTS       & CODEPOINTS & Unicode text by code point             \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}TEXT             & TEXT       & Unicode text by grapheme cluster       \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{CODEPOINTS\} & CODEPOINTS & Unicode text by code point             \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{TEXT\}       & TEXT       & Unicode text by grapheme cluster       \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{ISO8859-1\}  & Legacy     & ISO 8859-1 legacy text                 \\
\hline                                                     & CODEPOINTS & Unicode text by code point             \\
{\verb|`|}Ren\'{e}{\verb|`|}
                                                           & TEXT       & Unicode text by grapheme cluster (TBD) \\
\hline
{\verb|`|}Ren\'{e}{\verb|`|}C                              & CODEPOINTS & Unicode text by code point             \\
\hline
{\verb|`|}Ren\'{e}{\verb|`|}T                              & TEXT       & Unicode text by grapheme cluster       \\
\hline
\end{tabular}

An issue that must be resolved is whether string literals should have
binary fidelity or visual fidelity to the target code page, i.e.,
should the quoted source characters be given in the source code page
or in the target code page.  E.g., should the word "foo" in ISO 8859-1
and in EBCDIC be coded as
{\textquotesingle}foo{\textquotesingle}{ISO8859-1} and
{\textquotesingle}foo{\textquotesingle}{EBCDIC}, requiring that Rexx
translate the text, or should the second be entered with the EBCDIC
code points 86, 96, 96, making it harder to read and edit?

@@

\begin{itemize}
\item Use octet count, as at present. \\
This risks a program doing a seek within a UTF-8 sequence,
within a UCS-2/UCS-4 character,
within a UTF-16 surrogate pair,
or, for EGC access, withing an EGC.
Strategies to ameliorate this include
\begin{definition}
\item [UTF-8]
Read the octet at the specified offset and verify that either the high
bit is 0 or the two high order bits are 1.
\item [UCS-2]
Verify that the offset is a multiple of 2.
\item [UTF-16]
Verify that the offset is a multiple of 2.  Read two octets at the
specified offset and verify that is not a low surrogate (U+DC00-U+DFFF),
\item [UCS-4]
Verify that the offset is a multiple of 2.
Read four bytes and verify that they do not exceed U+10FFFF.
\end{definition}
\item Use character or EGC count \\
This is expensive
\item Define a cursor type. \\
This violates "Everything is a string" in classic Rexx.
\end{itemize}

@@

Case folding of symbols and caseless comparisons will use the \cite{UCD}.
\cref{issue:CF,issue:CComp,issue:CConv})
The rules for case folding should take into account mathematical usage.
\begin{itemize}
\item U+1D400 through U+1D7FF are semantically distinct from other letters.
\item Superscripts and subscripts have semantic significance.
\end{itemize}

@@

\section{Bibliography}

\subsection{GitHub ARB documents}
\renewcommand{\refname}{ARB}

\begin{thebibliography}{20}
\bibitem[New Types]{New Types}
\href{https://github.com/RexxLA/rexx-repository/edit/master/ARB/standards/work-in-progress/unicode/UnicodeTools/doc/string-types.md}%
{\textit{New types of strings}}

\bibitem[Requirements]{Req}
\href{https://github.com/users/RexxLA/projects/2/views/6}%
{\textit{Unicode Standard Development - requirements}}

\bibitem[Requirements Status]{ReqStat}
\href{https://github.com/users/RexxLA/projects/2/views/1}%
{\textit{Unicode Standard Development - requirements status}}

\bibitem[RXU]{RXU}
\href{https://github.com/RexxLA/rexx-repository/blob/master/ARB/standards/work-in-progress/unicode/UnicodeTools/rxu.md}%
{\textit{The RXU Rexx Preprocessor for Unicode}}

\end{thebibliography}

\subsection{Official documents}
\renewcommand{\refname}{References}

\begin{thebibliography}{20}
\bibitem[Glossary of Unicode Terms]{Glossary}
\href{https://unicode.org/glossary/}%
{\textit{Glossary of Unicode Terms}}

\bibitem[ISO/IEC 8859-1:1997 (E)]{ISO 8859-1}
\href{https://www.open-std.org/JTC1/sc2/wg3/docs/n411.pdf}%
{\textit{Final Text of DIS 8859-1, 8-bit single-byte coded graphic character sets-Part 1: Latin alphabet No.1}}

\bibitem[ISO/IEC 10646:2020]{ISO 10646}
\href{https://www.iso.org/standard/76835.html}%
{\textit{Information technology - Universal coded character set (UCS)}}

\bibitem[RFC 2781]{RFC 2781}
\href{https://datatracker.ietf.org/doc/rfc2781/}%
{\textit{UTF-16, a transformation format of ISO 10646}}

\bibitem[RFC 3629, STD 63]{RFC 3629}
\href{https://datatracker.ietf.org/doc/rfc3629/}%
{\textit{UTF-8, a transformation format of ISO 10646}}

\bibitem[RFC 5137, BCP 137]{RFC 5137}
\href{https://datatracker.ietf.org/doc/rfc5137/}%
{\textit{ASCII Escaping of Unicode Characters}}

\bibitem[RFC 5198]{RFC 5198}
\href{https://datatracker.ietf.org/doc/rfc5198/}%
{\textit{Unicode Format for Network Interchange}}

\bibitem[Unicode]{Unicode}
\href{https://www.unicode.org/standard/standard.html}%
{\textit{The Unicode Standard}}

\bibitem[Unicode]{Unicode3.9}
\href{https://www.unicode.org/versions/Unicode15.0.0/ch03.pdf#page=50}%
{\textit{3.9 Unicode Encoding Forms}}

\bibitem[Unicode Character Database]{UCD}
\href{https://www.unicode.org/ucd/}%
{\textit{Unicode Character Database (UCD)}}

\bibitem[Unicode CLDR Project]{CLDR}
\href{https://cldr.unicode.org/}%
{\textit{Unicode Common Locale Data Repository (CLDR)}}

\bibitem[Unicode{\textregistered} Standard Annex 29]{Annex29}
\href{https://www.unicode.org/reports/tr29/}%
{\textit{Unicode{\textregistered} Standard Annex 29\\ %
Unicode Text Segmentation}}

\bibitem[Unicode{\textregistered} Standard Annex 31]{Annex31}
\href{https://www.unicode.org/reports/tr31/}%
{\textit{Unicode{\textregistered} Standard Annex 31\\ %
Unicode Identifier and Pattern Syntax}}

\bibitem[Unicode{\textregistered} Standard Annex 42]{Annex42}
\href{https://www.unicode.org/reports/tr42/tr42-34.html}%
{\textit{Unicode{\textregistered} Standard Annex 42\\ %
Unicode Character Database in XML}}

\bibitem[Unicode{\textregistered} Standard Annex 44]{Annex44}
\href{https://www.unicode.org/reports/tr44/tr44-32.html}%
{\textit{Unicode{\textregistered} Standard Annex 44\\ %
Unicode Character Database}}

\end{thebibliography}

@@

The definitions given in \cite{Glossary} and \cite{Unicode} take
precedence over those given here. Quoted text is taken from those sources, or from IETF documents.
Except for definitions taken from official IETF and Unicode documents,
the nomenclature used here is illustrative rather than normative;
language design teams will formally define, e.g., method names, encoding of parameters.

\begin{definition}
\item [.BYTES] The string class for uninterprted octet strings.
\item [.CODEPOINTS] The string class for strings of Unicode code points.
\item [.Legacy] The string class for strings in legacy code pages.
May be merged with .BYTES.
\item [.TEXT] The string class for strings of Unicod extended grapheme clusters.
\item[ARB] RexxLA Architecture Review Board
\item [Bidi] "Abbreviation of bidirectional, in reference to mixed left-to-right and right-to-left text."
\item [BIF] Built In Function
\item [BIM] Built In Method
\item [Block] "A grouping of characters within the Unicode encoding
space used for organizing code charts. Each block is a uniquely named,
continuous, non-overlapping range of code points, containing a
multiple of 16 code points, and starting at a location that is a
multiple of 16. A block may contain unassigned code points, which are
reserved."
\item [BMP] Basic Multilingual Plane\index{BMP!definition}index{BMP!Basic Multilingual Plane!definition}
\\*
The first 64 Ki code points of Unicode, from U+0000 to U+FFFF.
\item [BOM] Byte Order Mark: U+FEFF, ZERO WIDTH NON-BREAKING SPACE (ZWNBSP)
\\*
Used as the first character to indicate byte order for UCS-2, UTF-16 and UCS-4;
\\*
Optional as the first character for UTF-8.
\item [encoded character] The smallest constituant of a Unicode string.
"The Unicode Standard does not define what is and is not a text
element in different processes; instead, it defines elements called
encoded characters. An encoded character is represented by a number
from 0 to 10FFFF\textsubscript{16} called a code point."
\item [EGC] Extended grapheme cluster
\item [GCGID] "Acronym for Graphic Character Global Identifier.
These are listed in the IBM document
Character Data Representation Architecture, Level 1, Registry SC09-1391."
See \url{https://www.ibm.com/downloads/cas/G01BQVRV}.
\item [grapheme cluster]
"A grapheme cluster consists of a base character followed by any
number of continuing characters, where acontinuing character may
include any nonspacing combining mark, certain spacing combining
marks, or a join control."
{\cite{Annex29}} defines two types of grapheme clusters;
"An extended grapheme cluster is the same as a legacy grapheme cluster,
with the addition of some other characters.
The continuing characters are extended to include all spacing combining marks,
such as the spacing (but dependent) vowel signs in Indic scripts."
\item [high surrogate] A code point in the range U+D800-U+DBFF, used
as the first half of a surrogate pair.
\item [IETF] \href{https://ietf.org}{Internet Engineering Task Force}
\item [introducer] The first octet in the UTF-8 encoding of a Unicode character beyond U+7F
\item [low surrogate] A code point in the range U+DC00-U+DFFF, used
as the second half of a surrogate pair.
\item [NFC]  Unicode
"Normalization Form C (NFC). A normalization form that erases any
canonical differences, and generally produces a composed result. For
example, a + umlaut is converted to \symbol{"00E4} in this form. This form most
closely matches legacy usage. The formal definition is D120 in Section
3.11, Normalization Forms."
\\*
the normalization endorsed by the IETF
\item [NFD] Unicode
"Normalization Form D (NFD). A normalization form that erases any
canonical differences, and produces a decomposed result. For example,
\symbol{"00E4} is converted to a + umlaut in this form. This form is most often
used in internal processing, such as in collation. The formal
definition is D118 in Section 3.11, Normalization Forms."

@@

>>>>

%Compile with latexmk -lualatex
%
% Changes
% Table of contents
% \underline{default}
% Split nomenclature section
% Link to 3.9 Unicode Encoding Forms
% Codepoints/EGCs option in support of charin()
% Correct and expand constants example
% Index
% stream - seek and update
% TUTOR
% Reference ARB GitHub issues
%
% \documentclass[draft]{article}
\documentclass[b4paper]{article}
\usepackage[
    backend=biber,%
    backref=true,%
    indexing=true,%
    sorting=anyt,%,
    style=alphabetic%
    ]%
    {biblatex}
\addbibresource{ARB.bib}
\addbibresource{FAQ.bib}
\addbibresource{IETF.bib}
\addbibresource{ISO.bib}
\addbibresource{Unicode.bib}
\usepackage{enumitem}
\usepackage{eqparbox}
\usepackage{expl3}
\usepackage{fontspec}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{ifthen}
\usepackage{imakeidx}
\usepackage{listings}         %Requires luatex engine to handle UTF-8
                              %Consider using minted, but ...
\usepackage{makeidx}
\usepackage{tocbibind}
\usepackage{upquote}
\usepackage{cleveref}

\makeindex[intoc]
\setmainfont{CMU Serif}
% texlive-cm-unicode or retrieve cm-unicode with MikTex Console

@@

\newcommand \fullcref [1]
   { \ifthenelse {\equal{\nameref{#1}}{}}
       {\cref{#1}}
       {\cref{#1} (\nameref{#1})}
   }

@@

\section{Background}
The Architecture Review board is investigating areas in which Rexx may be
in need of enhancements, some of which are listed in \cite{ARB:Issues}.
Many modern languages support Unicode encoded as
\href{https://datatracker.ietf.org/doc/rfc3629/}{UTF-8} (\cite{RFC:3629}),
and the \href{https://www.ietf.org/}{IETF}, in \cite{RFC:5198},
has mandated the use of UTF-8 with NFC in new protocols.
Rexx will need to support UTF-8 if it is not to become a backwater,
and that support should not break existing code.
Elaboration of this issue may be added to \cite{ARB:Issues} \cref{issue:MB,issue:UTF-8}.

\section{Scope}
This document only describes issues related to \cref{issue:EA,issue:MB}, expanded alphabet and
multi-octet representations of characters in Unicode \cref{issue:MB,issue:UTF-8}; it does not
address such issues as bidirectional text, nor does it address legacy
DBCS support. It presents suggested solutions for requirements that
are presented in other documents.  Also, most of the details are under
active discussion, so everything here is provisional. Information on Unicode may be found at \citetitle{Unicode} and \citetitle{UFAQs}.

@@

The definitions given in \cite{USTD:Gloss} and \cite{Unicode} take
precedence over those given here. Quoted text is taken from those sources.
Except for definitions taken from official IETF and Unicode documents,
the nomenclature used here is illustrative rather than normative;
language design teams will formally define, e.g., method names, encoding of parameters.
Most of the sections assume that there will be destinct string types
for extended grapheme clusters (\cref{issue:NT,issue:c2u}), Unicode code points (\cref{issue:NT,issue:c2u}),
legacy code pages (\cref{issue:Leg,issue:mLS,issue:iconv})
and raw octets (\cref{issue:ROS,issue:NT,issue:mO}). However, there has been discussion of including
metadata, in which case come of the classes might be merged.

@@

\begin{lstlisting}[caption={Non-ASCII upper casing in classic Rexx}]
    foo = 'Ren!\'{e}!'
    parse upper var foo bar
\end{lstlisting}

@@

\subsection{Numeric string conversion to characters}
\label{problem:numeric}
The definitions of binary and hexadecimal literals, and of the b2c()
and x2c() functions, in \cite{ANSI:J18}, depend on Config\_B2C. While the
definition"Translate Binary to a coded string." of Config\_B2C in ANSI
X3J18-1996 is rather vague, it is adequate when the character set is
restricted to an 8-bit character set.  The definition must be replaced
by one that is unamgiguous when applied to Unicode.
(\cref{issue:CUC}) There are several obvious options.

\begin{enumerate}
\item Treat every 32 bits as a Unicode scalar
\item Treat every 8 bits as part of a UTF-8 sequence
\item Treat every 8 bits as a character in a legacy code page.
\end{enumerate}

Each of these options has disadvantages.

The converse issue exists for Config\_C2B.

\subsection{Identifiers, labels and procedure names}
The current definition restricts symbols to ASCII characters, which
excludes letters used in many languages. If the rules are extended, in
accordance with \cite{UAX31}. to include Unicode letters and digits
beyond ASCII, the rules for equality of symbols must be addressed.
Are composed and decomposed strings identical? Are base characters
identical to their compatibility alterrnates? Are subscript and
superscript digits distinct from their ASCII counterparts?

When a numeric value is used as a label or procedure name, a precise
definition of the semantics is needed.  This is essentially the same
issue as \fullcref{problem:numeric} above.

@@

\subsection{PARSE}
The PARSE tempates currently support spliiting text based on a
restricted set of whitespace characters.  A decision must be made
whether to only support space and tab as whitespace for PARSE, or to
use one of the whitespace definitions in \cite{UAX31}.

@@

\subsection{Unassigned and invalid code points}
There has been some discussion of the need to detect
unassigned\footnote{Unsigned is not the same as noncharacter.}
code points. (\cref{issue:UICP,issue:ED})

@@

\subsection{Tools}
There are tools available that manipilate Rexx source code in various
ways, e.g., code completion, cross referencing, navigation, prettyprinting, syntax highligting.
These tools may not handle new syntax added in support of Unicode. Examples include
\begin{itemize}
\item The IBM Z\textregistered Open Editor \\
A Rexx Language Support plugin for Visual Studio Code
\item The \LaTeX\  package \href{https://www.ctan.org/pkg/listings}{\textbf{listings}}
\item The \LaTeX \ package \href{https://www.ctan.org/pkg/minted}{\textbf{minted}}
\item The Python package Pygments
\end{itemize}

\section{ARB GitHub Issues}
\label{ARBiss}

This section lists all issues in \cite{ARB:Issues}, whether or not they relate to Unicode.

\begin{issue}[label=I.\arabic*,ref=\fbox{I.\arabic*}]
\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730857}%
{Expanded Alphabets}}
\label{issue:EA}

@@

The RXU preprocessor accepts \textbf{OPTIONS DEFAULTSTRING} \textit{default}, where default can be one of \textbf{ BYTES},\textbf{ CODEPOINTS}, \textbf{TEXT} OR \textbf{NONE}.

@@

RXU uses BYTES for both octet strings and legacy strings.

@@

Assuming that the width is in in terms of octets
\begin{lstlisting}[
    caption={c2x() with width parameter},%
    linewidth=35em%
    ]
    foo = 'Ren!\'{e}!'
    bar = foo~UTF-8
    /* assumes that width parameter is in octets */
    say foo~c2x(2) /* Displays 00520065006E00E9  */
    say bar~c2x    /* Displays 52656EC3A9        */
\end{lstlisting}

There is an ongoing discussion as to whether an when to allow implicit coercions of type in, e.g.,
\begin{lstlisting}[
    caption={Implicit coercion},%
    linewidth=35em]%
    ]
    foo = .BYTES~new
    foo = 'Ren!\'{e}!'
    !\textellipsis!
    parse var foo ASCII !{\textquotesingle}\'{e}{\textquotesingle}! .
    say ASCII                     /* Displays Ren */
\end{lstlisting}

@@

\subsection{Constants}
Other languages allow specifying Unicode characters using either
the hexadecimal value of the code point or the assigned name of the code point,
(\cref{issue:CUC})
e.g., U+E9
(\cref{issue:NN})
might be coded as {\textbackslash}u\{E9\} or {\textbackslash}u[LATIN SMALL LETTER E WITH ACUTE].
The syntax used for such constants should be consisten across Rexx variants,
including rules for optional spaces between (hex) digits,
and should take into account the recommendations in \cite{RFC:5137}.
Implementations should use the machine readable data bases published by
the Unicode Consortium in order to ease migration to new versions of Unicode.

@@

\begin{tabular}{| l | l | l |}
\hline
Example & Type & Semantics \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}               & CODEPOINTS & Unicode text by code point             \\
                                                           & TEXT       & Unicode text by grapheme cluster (TBD) \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}R                & CODEPOINTS & Unicode text by code point             \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}CODEPOINTS       & CODEPOINTS & Unicode text by code point             \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}TEXT             & TEXT       & Unicode text by grapheme cluster       \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{CODEPOINTS\} & CODEPOINTS & Unicode text by code point             \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{TEXT\}       & TEXT       & Unicode text by grapheme cluster       \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{ISO8859-1\}  & Legacy     & ISO 8859-1 legacy text                 \\
\hline                                                     & CODEPOINTS & Unicode text by code point             \\
{\verb|`|}Ren\'{e}{\verb|`|}
                                                           & TEXT       & Unicode text by grapheme cluster (TBD) \\
\hline
{\verb|`|}Ren\'{e}{\verb|`|}C                              & CODEPOINTS & Unicode text by code point             \\
\hline
{\verb|`|}Ren\'{e}{\verb|`|}T                              & TEXT       & Unicode text by grapheme cluster       \\
\hline
\end{tabular}
\\
An issue that must be resolved is whether string literals should have
binary fidelity or visual fidelity to the target code page, i.e.,
should the quoted source characters be given in the source code page
or in the target code page.  E.g., should the word "foo" in ISO 8859-1
and in EBCDIC be coded as
{\textquotesingle}foo{\textquotesingle}{ISO8859-1} and
{\textquotesingle}foo{\textquotesingle}{EBCDIC}, requiring that Rexx
translate the text, or should the second be entered with the EBCDIC
code points 86, 96, 96, making it harder to read and edit?

Another issue is whether numeric constants used as labels and call
targets should be treated as strings of legacy code points, UTF-8
sequences or as strings of Unicode scalars.

RXU has five tpes of string literals:
\begin{definition}
\item['string'] The type is determined by OPTIONS.
\item['string'P] Unicode by code point
\item['string'T] Unicode by EGC
\item['space separated tokens'U] Unicode code points by alias, name or value
\item['string'Y] Legacy bYte string
\end{definition}

@@

\begin{itemize}
\item Use octet count, as at present. \\
This risks a program doing a seek within a UTF-8 sequence,
within a UCS-2/UCS-4 character,
within a UTF-16 surrogate pair,
or, for EGC access, within an EGC.
Strategies to ameliorate this include
\begin{definition}
\item [UTF-8]
Read the octet at the specified offset and verify that either the high
bit is 0 or the two high order bits are 1.
\item [UCS-2]
Verify that the offset is a multiple of 2.
\item [UTF-16]
Verify that the offset is a multiple of 2.  Read two octets at the
specified offset and verify that is not a low surrogate (U+DC00-U+DFFF),
\item [UCS-4]
Verify that the offset is a multiple of 2.
Read four bytes and verify that they do not exceed U+10FFFF.
\end{definition}
\item Use character or EGC count \\
This is expensive
\item Define a cursor type. \\
This violates "Everything is a string" in classic Rexx, but is reasonable for ooRexx.
\end{itemize}

@@

Case folding of symbols and caseless comparisons will use the \citetitle{UCD}.
\cref{issue:CF,issue:CComp,issue:CConv})
The rules for case folding should take into account mathematical usage.
\begin{itemize}
\item U+1D400 through U+1D7FF are semantically distinct from other letters.
\item Superscripts and subscripts have semantic significance.
\end{itemize}

@@

@@

The definitions given in \cite{USTD:Gloss} and \cite{Unicode} take
precedence over those given here. Quoted text is taken from those sources, or from IETF documents.
Except for definitions taken from official IETF and Unicode documents,
the nomenclature used here is illustrative rather than normative;
language design teams will formally define, e.g., method names, encoding of parameters.

\begin{definition}
\item [.BYTES] The string class for uninterprted octet strings.
\item [.CODEPOINTS] The string class for strings of Unicode code points.
\item [.Legacy] The string class for strings in legacy code pages.
May be merged with .BYTES.
\item [.TEXT] The string class for strings of Unicod extended grapheme clusters.
\item[ARB] RexxLA Architecture Review Board
\item [Bidi] "Abbreviation of bidirectional, in reference to mixed left-to-right and right-to-left text."
\item [BIF] Built In Function
\item [BIM] Built In Method
\item [Block] "A grouping of characters within the Unicode encoding
space used for organizing code charts. Each block is a uniquely named,
continuous, non-overlapping range of code points, containing a
multiple of 16 code points, and starting at a location that is a
multiple of 16. A block may contain unassigned code points, which are
reserved."
\item [BMP] Basic Multilingual Plane\index{BMP!definition}index{BMP!Basic Multilingual Plane!definition}
\\*
The first 64 Ki code points of Unicode, from U+0000 to U+FFFF.
\item [BOM] Byte Order Mark: U+FEFF, ZERO WIDTH NON-BREAKING SPACE (ZWNBSP)
\\*
Used as the first character to indicate byte order for UCS-2, UTF-16 and UCS-4;
\\*
Optional as the first character for UTF-8.
\item [encoded character] The smallest constituant of a Unicode string.
"The Unicode Standard does not define what is and is not a text
element in different processes; instead, it defines elements called
encoded characters. An encoded character is represented by a number
from 0 to 10FFFF\textsubscript{16} called a code point."
\item [EGC] Extended grapheme cluster
\item [GCGID] "Acronym for Graphic Character Global Identifier.
These are listed in the IBM document
Character Data Representation Architecture, Level 1, Registry SC09-1391."
See \url{https://www.ibm.com/downloads/cas/G01BQVRV}.
\item [grapheme cluster]
"A grapheme cluster consists of a base character followed by any
number of continuing characters, where acontinuing character may
include any nonspacing combining mark, certain spacing combining
marks, or a join control."
{\cite{UAX29}} defines two types of grapheme clusters;
"An extended grapheme cluster is the same as a legacy grapheme cluster,
with the addition of some other characters.
The continuing characters are extended to include all spacing combining marks,
such as the spacing (but dependent) vowel signs in Indic scripts."
\item [high surrogate] A code point in the range U+D800-U+DBFF, used
as the first half of a surrogate pair.
\item [IETF] \href{https://ietf.org}{Internet Engineering Task Force}
\item [introducer] The first octet in the UTF-8 encoding of a Unicode character beyond U+7F
\item [low surrogate] A code point in the range U+DC00-U+DFFF, used
as the second half of a surrogate pair.
\item [NFC]  Unicode
"Normalization Form C (NFC). A normalization form that erases any
canonical differences, and generally produces a composed result. For
example, a + umlaut is converted to \symbol{"00E4} in this form. This form most
closely matches legacy usage. The formal definition is D120 in Section
3.11, Normalization Forms."
\\*
the normalization endorsed by the IETF
\item [NFD] Unicode
"Normalization Form D (NFD). A normalization form that erases any
canonical differences, and produces a decomposed result. For example,
\symbol{"00E4} is converted to a + umlaut in this form. This form is most often
used in internal processing, such as in collation. The formal
definition is D118 in Section 3.11, Normalization Forms."

@@

\section{Bibliography}
This version of the document prints all included bibliography entries, even those note cited.
\nocite{*}

\printbibliography[%
    keyword=ARB,%
    title={GitHub ARB documents}%
    ]
\printbibliography[%
    keyword=FAQ,%
    title=FAQs%
    ]
\printbibliography[%
    keyword=standard,%
    notkeyword=FAQ,%
    title={Official documents}%
    ]
